---
title: "Analyzing expression data for hybrids relative to their parents"
author: 
  - name: Fabricio Almeida-Silva
    affiliation: 
    - VIB-UGent Center for Plant Systems Biology, Ghent, Belgium
    - Department of Plant Biotechnology and Bioinformatics, Ghent University, Ghent, Belgium
  - name: Lucas Prost-Boxoen
    affiliation: 
    - VIB-UGent Center for Plant Systems Biology, Ghent, Belgium
    - Department of Plant Biotechnology and Bioinformatics, Ghent University, Ghent, Belgium
  - name: Yves Van de Peer
    affiliation: 
    - VIB-UGent Center for Plant Systems Biology, Ghent, Belgium
    - Department of Plant Biotechnology and Bioinformatics, Ghent University, Ghent, Belgium
    - College of Horticulture, Academy for Advanced Interdisciplinary Studies, Nanjing Agricultural University, Nanjing, China
    - Center for Microbial Ecology and Genomics, Department of Biochemistry, Genetics and Microbiology, University of Pretoria, Pretoria, South Africa
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: yes
bibliography: vignette_bibliography.bib
vignette: >
  %\VignetteIndexEntry{Analyzing expression data for hybrids relative to their parents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

# Introduction

The formation of hybrids through the fusion of distinct genomes and subsequent 
genome duplication in cases of allopolyploidy represents a significant 
evolutionary event with complex effects on cellular biology, particularly gene expression. The impact of such genome mergings and duplications on 
transcription remain incompletely understood. To bridge this gap, we introduce __HybridExpress__, a comprehensive package designed to facilitate the 
comparative transcriptomic analysis of hybrids and their progenitor species.
__HybridExpress__ is tailored for RNA-Seq data derived from an 
'experimental trio': the hybrid organism and its two parental species. 
This package offers a suite of intuitive functions enabling researchers to 
perform differential expression analysis with ease, generate principal 
component analysis (PCA) plots for visualizing gene expression trends, 
categorize genes into 12 distinct expression pattern groups 
(as in @rapp2009genomic), and conduct in-depth functional analyses.
Acknowledging the potential variability in cell and transcriptome size 
across species and ploidy levels, __HybridExpress__ incorporates features for 
rigorous normalization of count data. Specifically, it allows for the 
integration of spike-in standards directly into the normalization process, 
ensuring accurate transcriptome size adjustments when these standards are 
present in the RNA-Seq count data (see full methodology in @coate2023beyond). 
By offering these capabilities, __HybridExpress__ provides a robust toolset for unraveling the intricate effects of genome doubling and merging on hybrid gene expression, paving the way for novel insights into the cellular biology of hybrid organisms.

# Installation

`r BiocStyle::Biocpkg("HybridExpress")` can be installed from Bioconductor
with the following code:

```{r installation, eval=FALSE}
if(!requireNamespace('BiocManager', quietly = TRUE))
  install.packages('BiocManager')

BiocManager::install("HybridExpress")
```

```{r load_package, message=FALSE}
# Load package after installation
library(HybridExpress)
```

# Data description





# Adding midparent expression values


1. Mean (default)
2. Sum
3. Weighted mean

# Exploratory data analyses



# Identifying differentially expressed genes between hybrids and their parents


# Expression-based gene classification



# FAQ {.unnumbered}

1. How do I create a `SummarizedExperiment` object?


# Session information {.unnumbered}

This document was created under the following conditions:

```{r, echo = FALSE}
sessioninfo::session_info()
```

# References {.unnumbered}







